---
title: "Differential Abundance" 
author: "Augustus Pendleton"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    highlight: default
    keep_md: yes
    theme: journal
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
editor_options: 
  chunk_output_type: console
---
<style>
pre code, pre, code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>


```{r setup, include=FALSE}
# For width of code chunks and scroll bar 
options(width=250)

options(getClass.msg=FALSE)

knitr::opts_chunk$set(eval = TRUE, 
                      echo = TRUE, 
                      include = TRUE,
                      warning = FALSE,
                      collapse = FALSE,
                      message = FALSE,
                      dpi=300, dev = "png",
                      engine = "R", # Chunks will always have R code, unless noted
                      error = TRUE,
                      fig.path="../figures/09_Differential_Abundance/",  
                      fig.align = "center") 

```

# Load packages 

```{r load-packages}
# Efficiently load packages 
pacman::p_load(phyloseq, ggpubr, tidyverse, ANCOMBC, microViz, ggrepel, combinat, install = FALSE)

knitr::write_bib(file = "data/09_differential_exports/packages.bib")

# load in functions and color preferences
source("code/R/plotting_aesthetics.R")

load("data/08_compositional_exports/full_abs_physeq.RData")

# This bit is necessary because otherwise ASV is treated as a taxonomic level above Kingdom, which causes problems.
releveled_physeq <- full_abs_physeq %>%
  speedyseq::relocate_tax_table(ASV, .after = last_col()) %>%
  speedyseq::mutate_tax_table(Species = ASV)

```


# Running ANCOM

I similarly ran many ANCOMs in preparing this, at different taxonomic levels and using different Comp_Groups (Depth-Month groups). This is the one we settled on for the paper. 

Cleaning up the output of ANCOM is a bit of a pain, so there's a lot of code for rearranging, filtering, and cleaning in this block. ANCOM also takes awhile to run - hence the running and storing as an object. 

```{r ancom-class-but-only-3-groups}

# class_ancom_3g_output <- ancombc2(data = releveled_physeq,
#                                 tax_level = "Class", # Test for each ASV
#                                 fix_formula = "Comp_Group_Hier", # Use Comp_Group_Hier to estimate diff. abundance
#                                 p_adj_method = "holm", # Adjust with Holm-Bonferroni correction; recommended by authors
#                                 pseudo_sens = TRUE, # Run sensitivity test to make sure taxa isn't sensitive to psuedo-count choice
#                                 prv_cut = 0.10, # Prevalence filter of 10%
#                                 group = "Comp_Group_Hier", # Use Comp_Group_Hier as groups when doing pairwise comparisons
#                                 struc_zero = TRUE, # Detect structural zeroes
#                                 alpha = 0.05, # Significance threshold of 0.05
#                                 n_cl = 6, # Use 6 threads
#                                 verbose = FALSE, # Don't print verbose output
#                                 global = TRUE, # Run a global test (sorta like an ANOVA to first find if a given ASV is sig diff between at least two groups)
#                                 pairwise = TRUE) # Run pairwise tests between groups (sorta like a post-hoc test like Tukey)
# 
# save(class_ancom_3g_output, file = "data/09_differential_exports/class_ancom_3g_output.RData")

load( "data/09_differential_exports/class_ancom_3g_output.RData")


clean_class <- class_ancom_3g_output$res_pair %>%
  select(taxon, starts_with("lfc"), starts_with("diff"), starts_with("passed_ss")) %>%
  pivot_longer(cols = !taxon, names_to = "metric", values_to = "value") %>%
  separate_wider_delim(cols = metric, delim = "_", names = c("variable", "Comparison"), too_many = "merge") %>%
  mutate(Comparison = str_remove(Comparison, "ss_")) %>%
  pivot_wider(id_cols = c("taxon","Comparison"), names_from = variable, values_from = value) %>%
  mutate(Comparison = str_remove(Comparison, "Comp_Group_Hier"),
         Comparison = str_replace(Comparison, "_Comp_Group_Hier", ";")) %>%
  separate_wider_delim(Comparison, delim = ";", names = c("Ref1", "Ref2"), too_few = "align_start") %>%
  mutate(Ref2 = ifelse(is.na(Ref2), "Deep", Ref2),
         Comparison = paste0(Ref1, " : ",Ref2)) %>% 
  dplyr::filter(diff == 1,
         passed == 1) %>%
  separate(taxon, into = c("Kingdom","Phylum","Class"), sep = "_") %>%
  dplyr::filter(Class != "NA") %>%
  select(Phylum, Class, Comparison, lfc)

clean_structural_zeros <- class_ancom_3g_output$zero_ind %>%
  dplyr::rename(One = 2, Two = 3, Three = 4) %>%
  filter(2|3|4) %>%
  separate(taxon, into = c("Kingdom","Phylum","Class"), sep = "_") %>%
  dplyr::filter(Class != "NA")
  

melted_3group <- psmelt(releveled_physeq)

```

So I had this fun idea for how to plot these data. I want to express both how strongly a given ASV is present in one group vs. another, while also showing its overall general abundance. To do this, I am defining each group (Deep, Shallow_May, and Shallow_September) as an axis along the unit circle (at 0, 2/3pi, and 4/3pi). Then, the mean abundance of each Class along each of those axes forms a triangle. I then calculate the centroid of that triangle, and plot it. These allows us to see when a given class trends two just one group, or sorta both groups, or isn't really that strongly different. It was a fun dive back into Trig and the unit circle.

```{r FIGURE-S9C, fig.width = 6.5, fig.height = 6.5}

centroid_data <- melted_3group %>%
  select(Class, Abundance, Comp_Group_Hier) %>%
  filter(Class %in% c(clean_class$Class, clean_structural_zeros$Class)) %>% # Select either differentially abundant or structural zero taxa
  group_by(Comp_Group_Hier, Class) %>%
  summarize(Mean_Abundance = mean(Abundance)) %>% # Find the mean absolute abundance of each class
  ungroup() %>%
  group_by(Class) %>%
  mutate(Rel_Abund = Mean_Abundance / sum(Mean_Abundance), # Find the mean relative abundance of each class within each group
         Theta = case_when(Comp_Group_Hier == "Deep"~.5 * pi, # This has to do with transforming points to polar space
                           Comp_Group_Hier == "Shallow_May"~(7/6) * pi,
                           Comp_Group_Hier == "Shallow_September" ~ (11/6) * pi),
         x = Rel_Abund * cos(Theta),
         y = Rel_Abund * sin(Theta)) %>%
  summarize(centr_x = mean(x),
            centr_y = mean(y))

# This next bit is to scale each class point by its maximum size (because they were ordinated by their mean abundance)

phy_and_abund <- melted_3group %>%
  select(Class, Abundance, Comp_Group_Hier, Phylum) %>%
  filter(Class %in%  c(clean_class$Class, clean_structural_zeros$Class)) %>%
  group_by(Phylum, Class) %>%
  summarize(Max_Abundance = max(Abundance)) %>%
  ungroup()

centroid_data %>% 
  left_join(phy_and_abund) %>%
  filter(Max_Abundance > 5000) %>%
  ggplot(aes(x = centr_x, y = centr_y)) + 
  annotate(geom = "segment", color = "grey80", 
           x = 0, y = 0, xend = .36 * cos(pi * c(.5, 7/6, 11/6)), yend = .36 * sin(pi * c(.5, 7/6, 11/6))) + 
  annotate(geom = "text", label = c("Deep","Shallow\nMay","Shallow\nSeptember"),
           x = c(.3,.37,.4) * cos(pi * c(.5, 7/6, 11/6)), y = c(0.37,0.37,0.4) * sin(pi * c(.5, 7/6, 11/6)),
           color = comp_three_colors[c(3,1,2)],
           fontface = "bold") + 
  geom_point(aes(size = Max_Abundance, color = Class)) + 
  geom_text_repel(aes(label = Class, color = Class), size = 3.3) + 
  coord_fixed(xlim = .4 * cos(pi * c(7/6, 11.5/6)), ylim = .4 * sin(pi * c(7/6, .5))) + 
  theme(axis.line = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank())+
  scale_size(guide = "none") + 
  scale_color_manual(values= class_colors) + 
  theme(legend.background = element_rect(color = "white"),
        legend.key.size = unit(2.9, "mm"),
        legend.title = element_blank(),
        panel.margin = unit(c(0,0,0,0), "mm"),
        legend.position = "none")  

```


## Session Info

```{r session_info}

sessioninfo::session_info()

```
