---
title: "Ecological Drivers (iCAMP)" 
author: "Augustus Pendleton"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    highlight: default
    keep_md: yes
    theme: journal
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
editor_options: 
  chunk_output_type: console
---
<style>
pre code, pre, code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>


```{r setup, include=FALSE}
# For width of code chunks and scroll bar 
options(width=250)

options(getClass.msg=FALSE)

knitr::opts_chunk$set(eval = TRUE, 
                      echo = TRUE, 
                      include = TRUE,
                      warning = FALSE,
                      collapse = FALSE,
                      message = FALSE,
                      dpi=300, dev = "png",
                      engine = "R", # Chunks will always have R code, unless noted
                      error = TRUE,
                      fig.path="../figures/12_Ecological_Drivers/",  
                      fig.align = "center") 

```

# Goals of this Document

In this analysis, we use the iCAMP package to infer the evolutionary processes governing community assembly both within our Comp groups and across them. This is the most computationally demanding section of our analysis - be aware that it takes a long time! Hence, many of these outputs are saved on disk, rather than being run afresh with each render. 

# Load packages 
```{r load-packages}

pacman::p_load(phyloseq, patchwork, iCAMP, tidyverse, ggside,  install = FALSE)

knitr::write_bib(file = "data/12_ecological_drivers/packages.bib")

# load in functions and color preferences
source("code/R/plotting_aesthetics.R")
```


# Load Data

```{r load_diversty_physeq}
load("data/08_compositional_exports/full_abs_physeq.RData")
```


# Prepare data objects

First, we format and export different objects from our physeq which we'll feed into iCAMP.

```{r exporting-objects}

# Environmental variables to test
env <- full_abs_physeq %>%
  sample_data %>%
  data.frame %>%
  dplyr::select(where(is.numeric)) %>%
  dplyr::select(NH4:temperature,good_oxygen,par) %>% 
  na.omit()

# We can only use samples with complete values
samples_w_env <- rownames(env)

env_physeq <- full_abs_physeq %>%
  prune_samples(samples_w_env, .) %>%
  prune_taxa(taxa_sums(.) > 0, .)

# We'll combine results using Comp_Group_Hier
pool_groups_env <- env_physeq %>%
  sample_data() %>%
  data.frame() %>%
  dplyr::select(Rep_ID, Comp_Group_Hier)

# Pull out our OTU table
asv_mat_env <- env_physeq %>%
  otu_table %>%
  as.matrix()

# Pull out our tree
tree_env <- env_physeq %>%
  phy_tree()

# Pull out and clean-up our tax table
tax_table_raw <- tax_table(env_physeq) %>% as.data.frame

tax_for_icamp <- tax_table_raw[,2:7]

# Set the working directory for our phylogenetic dist matrix

icamp_wd_env <- "data/12_ecological_drivers/pd_wd_env/"

# Make sure our Comp-Group_Hier groups match our OTU table
pools_env <- pool_groups_env$Comp_Group_Hier[match(row.names(asv_mat_env), pool_groups_env$Rep_ID)] %>%
  data.frame() %>%
  dplyr::rename(Comp_Group_Hier = 1)

row.names(pools_env) <- row.names(asv_mat_env)

```


# Phylogenetic dist matrix

This is an important step where we create a distance matrix which tracks the phylogenetic distance between each ASV.
```{r calculate-phylogenetic-dist-matrix, eval = FALSE}

pd.big=iCAMP::pdist.big(tree = tree_env, 
                        wd=icamp_wd_env, 
                        nworker = 50)

save(pd.big, file = "data/12_ecological_drivers/pd.big.RData")

```


# Environmental dist matrix

In this step, we create a dist matrix using our environmental variables, comparing each sample to every other sample. This will then be used later to correlate changes in our abundance within our bins to changes in environmental conditions. 

```{r calculate-environmental-dist-matrix, eval = FALSE}

niche.dif<-iCAMP::dniche(env = env,
                        comm = asv_mat_env,
                        method = "niche.value",
                        nworker = 30,
                        out.dist=FALSE,
                        bigmemo=TRUE,
                        nd.wd="data/12_ecological_drivers/output_env")

save(niche.dif, file = "data/12_ecological_drivers/niche.dif.RData")

```


# Optimizing our phylogenetic binning

One of the biggest choices we need to make when running iCAMP is how we define our phylogenetic bins. We can modulate this across two parameters:

1. The ds = the max phylogenetic distance at which to test phylogenetic signal strength

2. bin.size.limit = the minimum number of ASVs in each bin; bins smaller than this will get merged

We also test using both the RowSums and the RowMeans to aggregate the data across our environmental variables.

```{r finding-best-binning, eval = FALSE}

load("data/12_ecological_drivers/pd.big.RData")
load("data/12_ecological_drivers/niche.dif.RData")

dss <- c(0.1, 0.3, 0.5)
bin.size.limits <- c(24, 48)

combos <- expand_grid(dss, bin.size.limits)

test_binning <- function(ds, bin.size.limit){
    phylobin=taxa.binphy.big(tree = tree_env, 
                         pd.desc = pd.big$pd.file,
                         pd.spname = pd.big$tip.label,
                         pd.wd = pd.big$pd.wd, 
                         ds = ds, 
                         bin.size.limit = bin.size.limit,
                         nworker = 50)
    
    sp.bin=phylobin$sp.bin[,3,drop=FALSE]
    
    sp.ra=colMeans(asv_mat_env/rowSums(asv_mat_env))
    
    binps=ps.bin(sp.bin = sp.bin,
                    sp.ra = sp.ra,
                    pd.desc = pd.big$pd.file, 
                    pd.spname = pd.big$tip.label,
                    pd.wd = pd.big$pd.wd,
                    nd.list = niche.dif$nd,
                    nd.spname = niche.dif$names,
                    ndbig.wd = niche.dif$nd.wd,
                    cor.method = "pearson",
                    r.cut = 0.1, 
                    p.cut = 0.05
                    )
    sum_summaries <- binps$Index %>%
      dplyr::select(NH4.pearson:par.pearson) %>%
      rowSums
    
    mean_summaries <- binps$Index %>%
      dplyr::select(NH4.pearson:par.pearson) %>%
      rowMeans
    
    summaries <- data.frame(Summary = rep(c("Sum","Mean"), each = 4),
                            tested_ds = ds,
                            tested_bin.size.limit = bin.size.limit,
                            Metric = rep(binps$Index$index, 2),
                            Value = c(sum_summaries, mean_summaries)
               )
    
    return(summaries)
    
}


binning_tests <- map2(
  combos$dss,
  combos$bin.size.limits,
  possibly(\(x,y)test_binning(ds = x, bin.size.limit = y),
           otherwise = "ERRORED")
  )


final_results <- list_rbind(binning_tests)

save(final_results, file = "data/12_ecological_drivers/final_results.RData")

```


```{r plotting-ds-and-binlimits}

load("data/12_ecological_drivers/final_results.RData")

final_results %>%
  filter(Metric %in% c("RAsig.adj","MeanR"),
         Summary == "Sum") %>%
  ggplot(aes(x = tested_ds, 
             y = Value,
             color = factor(tested_bin.size.limit))) + 
  geom_line() + 
  geom_point() + 
  facet_wrap(~Metric,
             scales = "free_y") +
  labs(color = "bin.size.limit", x ="ds") + 
  theme_classic(base_size = 16)
```

# Run iCAMP

```{r icamp-run, eval = FALSE}

icamp_results_ds5_bsl24 <- icamp.big(comm = asv_mat_env,
                           tree = tree_env,
                           pd.desc = pd.big2$pd.file, 
                           pd.spname=pd.big2$tip.label,
                           pd.wd = pd.big2$pd.wd,
                           rand = 1000,
                           prefix = "icamp_ds5_bsl24_",
                           ds = .5,
                           bin.size.limit = 24,
                           nworker = 60,
                           detail.save = TRUE,
                           qp.save = TRUE,
                           detail.null = TRUE,
                           output.wd = "data/12_ecological_drivers/output_ds5_bsl24",
                           omit.option = "no",
                           taxo.metric = "bray",
                           sig.index = "Confidence",
                           ses.cut = 1.96, 
                           rc.cut = 0.95, 
                           conf.cut=0.975,
                           transform.method = NULL)

save(icamp_results_ds5_bsl24, file = "data/12_ecological_drivers/icamp_results_ds5_bsl24.RData")

```

# Calculating bin-level statistics

Now, we feed our icamp results into a function that will do several things at once, namely:

1. Summarize the dominant processes across our pools
2. Summarize which bins are most important for each sample's turnover
3. Summarize which processes are most important in each bin for each sample's turnover



```{r bin-level-stats, eval = FALSE}

icamp_bin_ds5_bsl24 <- icamp.bins(icamp.detail = icamp_results_ds5_bsl24$detail,
                        treat = pools_env,
                        clas=tax_for_icamp,
                        silent=FALSE, 
                        boot = TRUE,
                        rand.time = 1000,
                        between.group = TRUE)


save(icamp_bin_ds5_bsl24,
     file = "data/12_ecological_drivers/icamp_bin_ds5_bsl24.rda")
```

## Analysis

Now let's analyze some of these results

```{r reload-large-objects}
load("data/12_ecological_drivers/icamp_bin_ds5_bsl24.rda")
```

## Figure 3A/B

In this figure, we make treemap plots for the relative contribution of each class to each process within each group. We also scale the area of each process by its relative importance. There isn't a very good way to do this in R. So what I've done is exported each of the mini treemaps as its own separate plot into a folder call tree_panels, and then scaled and arranged them in Illustrator to produce Figure 3A.

```{r bars-with-taxa, fig.height = 6, fig.width=8}

icamp_bin <- icamp_bin_ds5_bsl24

bptk <- icamp_bin$BPtk

bin_classes <- icamp_bin$Bin.TopClass %>% 
  dplyr::select(Bin, Class.maxNamed) %>%
  mutate(Bin = tolower(Bin))


labeled_bptk <- bptk %>%
  pivot_longer(bin1:bin130, names_to = "Bin", values_to = "Contribution") %>%
  left_join(bin_classes) %>%
  dplyr::rename(Class = Class.maxNamed)
  
important_class <- labeled_bptk %>%
  group_by(Class) %>%
  summarize(max_cont = max(Contribution)) %>%
  filter(max_cont > 0.01) %>%
  pull(Class)
```

Here, we export panels for across-group comparisons, which together create Figure 3B

```{r scaling-stained-glass}
tree_plots <- labeled_bptk %>%
  mutate(Class = ifelse(Class %in% important_class, Class, "Rare")) %>%
  filter(!(Group %in% c("Deep","Shallow_September","Shallow_May")),
         Process %in% c("HoS","DR","DL")) %>%
  group_by(Group, Process, Class) %>%
  summarize(Contribution = sum(Contribution)) %>%
  mutate(Process = case_match(Process,
                              "DL" ~ "Dispersal Limitation",
                              "HoS" ~ "Homogenizing Selection",
                              "DR" ~ "Drift")) %>%
  ungroup() %>%
  nest_by(Process, Group) %>%
  mutate(plots = 
           list(ggplot(data = data, aes(area = Contribution, fill = Class)) + 
  treemapify::geom_treemap(start = "topleft") + 
  scale_fill_manual(values = class_colors, guide = "none") +
    theme_void()
           )
  )

# These are the panels which go into Figure 3A
for(i in 1:nrow(tree_plots)){
  ggsave(tree_plots$plots[[i]], 
         filename = paste("figures/12_Ecological_Drivers/tree_panels/",
                          paste(tree_plots$Process[i], tree_plots$Group[i], sep = "_"),
         ".png"),
         width = 1, height = 1, units = "in", create.dir = TRUE)
}

# Finding average heights

 labeled_bptk %>%
  mutate(Class = ifelse(Class %in% important_class, Class, "Rare")) %>%
  filter(!(Group %in% c("Deep","Shallow_September","Shallow_May")),
         Process %in% c("HoS","DR","DL")) %>%
  group_by(Group, Process) %>%
  summarize(Contribution = sum(Contribution)) %>%
  mutate(Process = case_match(Process,
                              "DL" ~ "Dispersal Limitation",
                              "HoS" ~ "Homogenizing Selection",
                              "DR" ~ "Drift")) %>%
   ungroup() %>%
   mutate(Ratio = Contribution / max(Contribution),
          Side_Adjustment = sqrt(Ratio))

```

Here, we export panels for within-group turnovers, which together create Figure 3A

```{r scaling-stained-glass-within-group}
tree_plots <- labeled_bptk %>%
  mutate(Class = ifelse(Class %in% important_class, Class, "Rare")) %>%
  filter((Group %in% c("Deep","Shallow_September","Shallow_May")),
         Process %in% c("HoS","DR","DL")) %>%
  group_by(Group, Process, Class) %>%
  summarize(Contribution = sum(Contribution)) %>%
  mutate(Process = case_match(Process,
                              "DL" ~ "Dispersal Limitation",
                              "HoS" ~ "Homogenizing Selection",
                              "DR" ~ "Drift")) %>%
  ungroup() %>%
  nest_by(Process, Group) %>%
  mutate(plots = 
           list(ggplot(data = data, aes(area = Contribution, fill = Class)) + 
  treemapify::geom_treemap(start = "topleft") + 
  scale_fill_manual(values = class_colors, guide = "none") +
    theme_void()
           )
  )

for(i in 1:nrow(tree_plots)){
  ggsave(tree_plots$plots[[i]], 
         filename = paste("figures/12_Ecological_Drivers/tree_panels_within_group/",
                          paste(tree_plots$Process[i], tree_plots$Group[i], sep = "_"),
         ".png"),
         width = 1, height = 1, units = "in", create.dir = TRUE)
}

# Finding average heights

 labeled_bptk %>%
  mutate(Class = ifelse(Class %in% important_class, Class, "Rare")) %>%
  filter(
         Process %in% c("HoS","DR","DL")) %>%
  group_by(Group, Process) %>%
  summarize(Contribution = sum(Contribution)) %>%
  mutate(Process = case_match(Process,
                              "DL" ~ "Dispersal Limitation",
                              "HoS" ~ "Homogenizing Selection",
                              "DR" ~ "Drift")) %>%
   ungroup() %>%
   mutate(Ratio = Contribution / max(Contribution),
          Side_Adjustment = sqrt(Ratio)) 


```

## Figure 3B: Relating bin abundance/variation to assembly process

In this section, I look at the overall abundance of each bin, and its coefficient of variation, and relate that to the dominant process which influences assembly within that bin. 

```{r calculating-bin-abundances}
Ptk <- icamp_bin$Ptk 

dps <- Ptk %>% 
  filter(Index == "DominantProcess") %>%
  pivot_longer(bin1:bin130, names_to = "Bin", values_to = "DominantProcess") %>% 
  select(Group : DominantProcess) %>%
  group_by(Bin) %>%
  count(DominantProcess)  %>% 
  ungroup()

bin_dom_procsses <- dps %>%
  group_by(Bin) %>%
  slice_max(n = 1, order_by = n) %>%
  mutate(Vals = n()) %>%
  ungroup()
  
bins_w_one_dom <- bin_dom_procsses %>%
  filter(Vals == 1) %>% 
  select(Bin, DominantProcess)

all_bin_process <- bin_dom_procsses %>%
  filter(Vals != 1) %>% 
  pivot_wider(names_from = DominantProcess, values_from = DominantProcess) %>%
  unite(DL:HeS, col = "DominantProcess", sep = "/", na.rm = TRUE) %>%
  select(Bin, DominantProcess) %>%
  rbind(bins_w_one_dom)
```

Now, let's calculate the average and maximum total abundance of each bin

```{r abund_prev_process_abs}
bin_assignments <- icamp_bin$Class.Bin %>%
  select(Bin) %>%
  rownames_to_column("ASV") %>%
  mutate(Bin = tolower(Bin))

melted_asv <- full_abs_physeq %>%
  psmelt() %>%
  select(ASV, Abundance, Rep_ID)

bin_sample_abunds <- melted_asv %>%
  left_join(bin_assignments) %>% 
  filter(!is.na(Bin)) %>%
  group_by(Bin, Rep_ID) %>%
  summarize(bin_sample_abund = sum(Abundance)) %>%
  ungroup()

bin_summarized_abunds <- bin_sample_abunds %>%
  group_by(Bin) %>%
  summarize(max_abund = max(bin_sample_abund),
            mean_abund = mean(bin_sample_abund),
            median_abund = median(bin_sample_abund),
            variance = sd(bin_sample_abund) / mean_abund)
```


```{r FIGURE-3C, fig.width = 3.5, fig.height = 3.5}
bin_summarized_abunds %>%
  left_join(all_bin_process) %>%
  filter(DominantProcess %in% c("DL", "DL/DR","DR","HoS")) %>%
  ggplot(aes(x = variance, y = max_abund, color = DominantProcess, fill = DominantProcess)) + 
  geom_point(size = 1.5, alpha = 0.7) + 
  geom_ysidedensity(alpha = 0.2, show.legend = FALSE) + 
  scale_x_continuous(transform = "log10", labels = scales::label_comma()) +
  scale_y_continuous(transform = "log10", labels = scales::label_comma()) +
  guides(x = guide_axis(check.overlap = TRUE)) + 
  scale_color_manual(values = process_colors,
                     breaks = c("HoS","DR","DL/DR","DL"),
                     labels = c("Homogenizing\nSelection",
                                "Drift",
                                "Dispersal Limitation/\nDrift",
                                "Dispersal Limitation")) + 
  scale_fill_manual(values = process_colors,
                    breaks = c("HoS","DR","DL/DR","DL"),
                     labels = c("Homogenizing\nSelection",
                                "Drift",
                                "Dispersal Limitation/\nDrift",
                                "Dispersal Limitation")) + 
  labs(x = "Coefficient of Variation",
       y = "Max Abundance",
       fill = "Dominant Process",
       color = "Dominant Process") + 
  theme(legend.position = "inside",
        legend.position.inside = c(0.6, 0.85),
        ggside.panel.scale = 0.2) + 
  scale_ysidex_continuous(labels = NULL) + 
  theme(axis.title = element_text(size = 9),
        axis.text = element_text(size = 8),
        strip.text = element_text(size = 8),
        legend.text= element_text(size = 7,
                                  margin = margin(0)),
        legend.title = element_text(size = 9),
        legend.key.size = unit(.5, "cm"),
        legend.box.background = element_blank())

filt_bin <- bin_summarized_abunds %>%
  left_join(all_bin_process) %>%
  filter(DominantProcess %in% c("DL", "DL/DR","DR","HoS"))

cor.test(filt_bin$max_abund,
       filt_bin$variance,
       method = "spearman")

```

